---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
This document contains the completed work for peer assignment #1 for the Coursera Reproducable Research class.

## Loading and preprocessing the data
1.Downloads data for Assessment and extrgetwacts it
```{r}  
 zip.url <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip'
 zip.file <- 'dataset.zip'
 download.file(zip.url, destfile = zip.file)
 download.data<-unzip(zip.file)
```
The variables included in this dataset are:

- steps: Number of steps taking in a 5-minute interval (missing values are coded as NA )
- date: The date on which the measurement was taken in YYYY-MM-DD format
- interval: Identifier for the 5-minute interval in which measurement was taken

2.Read the data into a data.frame 
```{r}
 data <- read.csv(download.data)
 data$date <- as.POSIXct(data$date) # set the dates to POSIXct
 Sys.setlocale("LC_TIME","English")
```


## What is mean total number of steps taken per day?

##Histogram, Mean & Median

1.Calculate the total number of steps taken per day
first I will summarize the data by day and transform it in table:
```{r}
library(plyr)
steps_per_day <- aggregate(data$steps, by = list(data$date), sum, na.rm=TRUE) 
names(steps_per_day) <- c("Date", "total_steps")
```

2.Make a histogram of the total number of steps taken each day
```{r}
hist(steps_per_day$total_steps,
             breaks = 20,
             main = 'Number of Steps Taken Per Day',
             xlab = 'Total Number of Steps', col = 'dark green',
             cex.main = .9)
```

3.Calculate and report the mean and median of the total number of steps taken per day
```{r}
mean_value = mean(steps_per_day$total_steps)
median_value = median(steps_per_day$total_steps)
```
The mean number of steps each day is `r mean_value`

The median number of steps each day is `r median_value`

##What is the average daily activity pattern?

1.Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
intsteps <- aggregate(data$steps, by = list(data$interval), mean, na.rm=TRUE)
intstepsmed <- aggregate(data$steps, by = list(data$interval), median, na.rm=TRUE)

intsteps <- cbind(intsteps[], intstepsmed$x)
```

Tidy the df names and round the numbers
```{r}
library(ggplot2)
names(intsteps) = c("interval","mean.steps", "median.steps")
intsteps$mean.steps <- round(intsteps$mean.steps)
intsteps$median.steps <- round(intsteps$median.steps)

ggplot(intsteps, aes(x = interval, y = mean.steps)) + geom_line()
```


2.Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
most.steps <- intsteps$interval[intsteps$mean.steps == max(intsteps$mean.steps)]
```

The interval with the most steps each day (on average is) : `r most.steps`


## Imputing missing values
Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

1.Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
Find the NAs
```{r}
na.steps <- subset(data, is.na(steps))
num.NAs <-length(na.steps$steps)
```
There are `r num.NAs` intervals with NA

2.Create a new dataset that is equal to the original dataset but with the missing data filled in.
Replace the NAs with the median number of steps for that period
```{r}
nstps <- data.frame(date=data$date[is.na(data$steps)], interval = data$interval[is.na(data$steps)], steps=intsteps[match(intsteps$interval, data$interval[is.na(data$steps)]),3])
```
Remove the NA's from the period
```{r}
data2<- subset(data, !is.na(steps))
```
Append the median steps to the Activity DF

```{r}
data2 <- rbind(data2, nstps)
```
3.Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 
sum the number of steps each day into the dailysteps2 DF and get the mean and median 
```{r}
steps_per_day2 <- aggregate(data2$steps, by = list(data2$date), sum, na.rm=TRUE)
names(steps_per_day2) <- c("Date", "total_steps")

qplot(total_steps, data = steps_per_day2, geom="histogram", xlab = "Daily Number of Steps", binwidth = 300)

```



```{r}
mean_value2 <- round(mean(steps_per_day2$total_steps),0)
median_value2 <- round(median(steps_per_day$total_steps), 0)
```
THe new mean number of steps is `r mean_value2` this is close to the mean from the data with NAs of `r mean_value`. 
The new median number of steps is `r median_value2` this is close to the median from the data with NAs of `r median_value`.
There is little impact to the estimated number of steps a day from using the median for the time interval to replace the missing data.


## Are there differences in activity patterns between weekdays and weekends?
1.Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
Add the Weekday/weekend identifier
```{r}
data2$week <- ifelse(weekdays(data2$date) == "Saturday" | weekdays(data2$date) == "Sunday" ,"weekend","weekday")
```

2.Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 
Df of the mean and median number of steps taken, averaged across all days (y-axis)
```{r}

intsteps2 <- aggregate(data2$steps, by = list(data2$week, data2$interval), mean, na.rm=TRUE)
intstepsmed2 <- aggregate(data2$steps, by = list(data2$week, data2$interval), median, na.rm=TRUE)

intsteps2 <- cbind(intsteps2[], intstepsmed2$x)

```

Tidy the df names and round the numbers
```{r}
names(intsteps2) = c("weekday", "interval","mean.steps", "median.steps")
intsteps2$mean.steps <- round(intsteps2$mean.steps)
intsteps2$median.steps <- round(intsteps2$median.steps)

ggplot(intsteps2, aes(x = interval, y = mean.steps)) + ylab("Number of Steps") + geom_line() + facet_grid(weekday~.)


As can be seen from the plots above the user appears to start walking around later at weekends. Once they are awake they tend to take more steps and to be walking around later in the day.